---
title: "Code for assignment 1"
author: "Dora, Rebecca, Jana, Asger, Sophia"
date: "10/10/19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading packages
pacman::p_load(tidyverse,janitor, stringr)

# Setting working directory
setwd("~/GitHub/github-files/Assignment-1")



## Reading the csv files with the training data sets, and saving them as tibbles
# Demographic data
demog_train = as_tibble(read.csv("demo_train.csv", sep = ","))

# Length of utterance data
LU_train = as_tibble(read.csv("LU_train.csv", sep = ","))

# Word data
token_train = as_tibble(read.csv("token_train.csv", sep = ","))
```



2a. Identify which variable names do not match (that is are spelled differently) and find a way to transform variable names.
Pay particular attention to the variables indicating participant and visit.

```{r}



```


2b. Find a way to homogeneize the way "visit" is reported (visit1 vs. 1).

```{r}
# The str_extract function is applied to the VISIT column of the word data and utterance data, in order to extract the number of visit. Those are then made into integers
word_data$VISIT <- word_data$VISIT %>% str_extract("\\d") %>% as.integer()

utt_data$VISIT <- utt_data$VISIT %>% str_extract("\\d") %>% as.integer()


```


2c. We also need to make a small adjustment to the content of the Child.ID coloumn in the demographic data. Within this column, names that are not abbreviations do not end with "." (i.e. Adam), which is the case in the other two data sets (i.e. Adam.). If The content of the two variables isn't identical the rows will not be merged.

```{r}
# Punctuation is removed from the subject column of the three data sets
word_data$SUBJ <- word_data$SUBJ %>% str_replace_all("[:punct:]", "")
utt_data$SUBJ <- utt_data$SUBJ %>% str_replace_all("[:punct:]", "")
demo_data$Child.ID <- demo_data$Child.ID %>% str_replace_all("[:punct:]", "")
```


2d. Now that the nitty gritty details of the different data sets are fixed, we want to make a subset of each data set only containig the variables that we wish to use in the final data set.
For this we use the tidyverse package dplyr, which contains the function select(). 

```{r}
#selecting the needed columns from each data set, making new dataframes with just these columns
demo_data_selected <- demo_data %>% select(Child.ID, Visit, Diagnosis, Ethnicity, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization)
word_data_selected <- word_data %>% select(SUBJ, VISIT, types_MOT, types_CHI, tokens_MOT, tokens_CHI)
utt_data_selected <- utt_data %>% select(SUBJ, VISIT, MOT_MLU, CHI_MLU)

#checking that they are all dataframes
class(demo_data_selected)
class(word_data_selected)
class(utt_data_selected)

#chanding the name of the column "Child.ID" into "SUBJ" and "Visit" into "VISIT"
colnames(demo_data_selected)[1] <- "SUBJ"
colnames(demo_data_selected)[2] <- "VISIT"

#another longer way to do the same thing:
#names(demo_data_selected)[names(demo_data_selected) == "Child.ID"] <- "SUBJ"
#names(demo_data_selected)[names(demo_data_selected) == "Visit"] <- "VISIT"
```

2e. Finally we are ready to merge all the data sets into just one. 

Some things to pay attention to:
* make sure to check that the merge has included all relevant data (e.g. by comparing the number of rows)
* make sure to understand whether (and if so why) there are NAs in the dataset (e.g. some measures were not taken at all visits, some recordings were lost or permission to use was withdrawn)

```{r}
merge_word_utt <- merge(word_data_selected, utt_data_selected)
merge_all_data <- merge(merge_word_utt, demo_data_selected)

#checking how many subject had less visits than 6
#for the demo dataframe (the bad boi)
group_demo <- demo_data_selected %>% group_by(SUBJ) %>% summarise(VISIT=n())
filter_demo <- group_demo %>% filter(group_demo$VISIT<6)

#for the word dataframe (which will be the same as the utt dataframe)
group_word <- word_data_selected %>% group_by(SUBJ) %>% summarise(VISIT=n())
filter_word <- group_word %>% filter(group_word$VISIT<6)

#for the final merged dataframe - which has only the subjects that are the same for all three dataframes
group_all <- merge_all_data %>% group_by(SUBJ) %>% summarise(VISIT=n())
filter_all <- group_all %>%  filter(group_all$VISIT<6)

#cbind_word_utt <- cbind(word_data_selected, utt_data_selected)
#cbind_all <- cbind(cbind_word_utt, demo_data_selected)

```


2f. Only using clinical measures from Visit 1
In order for our models to be useful, we want to minimize the need to actually test children as they develop. In other words, we would like to be able to understand and predict the children's linguistic development after only having tested them once. Therefore we need to make sure that our ADOS, MullenRaw, ExpressiveLangRaw and Socialization variables are reporting (for all visits) only the scores from visit 1.

A possible way to do so:
* create a new dataset with only visit 1, child id and the 4 relevant clinical variables to be merged with the old dataset
* rename the clinical variables (e.g. ADOS to ADOS1) and remove the visit (so that the new clinical variables are reported for all 6 visits)
* merge the new dataset with the old



```{r}
#new dataframe with only visit 1 data and from these 4 columns: ADOS, MullenRaw, ExpressiveLangRaw, Socialization (plus subject name)
new_df <- merge_all_data %>% filter(VISIT==1) %>% select(SUBJ, ADOS, MullenRaw, ExpressiveLangRaw, Socialization)

#renaming column names
colnames(new_df)[2] <- "ADOS1"
colnames(new_df)[3] <- "MullenRaw1"
colnames(new_df)[4] <- "ExpressiveLangRaw1"
colnames(new_df)[5] <- "Socialization1"

#merge it with the old one
new_merge <- merge(new_df, merge_all_data)
```

2g. Final touches

Now we want to
* anonymize our participants (they are real children!). 
* make sure the variables have sensible values. E.g. right now gender is marked 1 and 2, but in two weeks you will not be able to remember, which gender were connected to which number, so change the values from 1 and 2 to F and M in the gender variable. For the same reason, you should also change the values of Diagnosis from A and B to ASD (autism spectrum disorder) and TD (typically developing). Tip: Try taking a look at ifelse(), or google "how to rename levels in R".
* Save the data set using into a csv file. Hint: look into write.csv()

```{r}
#making gender into numerical instead of integer to be able to change it into M and F
new_merge$Gender <- as.numeric(new_merge$Gender)

new_merge$Gender[new_merge$Gender=="1"] <- "M"
new_merge$Gender[new_merge$Gender=="2"] <- "F"

#making the subject names (characters) into numbers. First change the string into factor
new_merge$SUBJ <- as.factor(new_merge$SUBJ)

levels(new_merge$SUBJ)[1:61] <- 1:61

#making it back into integer
new_merge$SUBJ <- as.integer(new_merge$SUBJ)

#A = 1 = ASD, B = 2 = TD
new_merge$Diagnosis <- as.numeric(new_merge$Diagnosis)
new_merge$Diagnosis[new_merge$Diagnosis=="1"] <- "ASD"
new_merge$Diagnosis[new_merge$Diagnosis=="2"] <- "TD"

write.csv(new_merge, file="dataset.csv")
```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
